---
title: "What Drives Healthcare Costs? A Statistical Analysis for Public Health Policy"
author: "Michael Matyevich"
date: "2026-04-10"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
library(tidyverse)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(xgboost)
library(Metrics)
library(pdp)
library(corrplot)
library(tidyr)
library(broom)
library(ggfortify)
library(scales)
library(here)

med_df <- read.csv(here("data", "insurance.csv")) |>
  mutate(
    sex    = as.factor(sex),
    smoker = as.factor(smoker),
    region = as.factor(region),
    bmi_cat = cut(
      bmi,
      breaks = c(0, 18.5, 25, 30, Inf),
      labels = c("Underweight", "Normal", "Overweight", "Obese"),
      right  = FALSE
    )
  )
```

---

## Background

Healthcare costs are rising in most developed countries, placing increasing strain on public health
systems and insurers alike. Understanding *which patient characteristics most strongly predict
higher costs* is critical for designing effective public health interventions — particularly when
resources are limited and policy decisions need to be evidence-based.

This analysis uses a dataset of 1,338 US insurance beneficiaries to answer three research questions.

---

## Research Questions

**RQ1 — What patient characteristics are most strongly associated with higher insurance charges?**

We expect age, BMI, and smoking status to be important based on the clinical literature, but the
relative magnitude of each, and whether they *interact* with one another, is not known in advance.
Understanding this helps prioritise which risk factors are most impactful to address.

**RQ2 — Does the effect of BMI on charges depend on whether a person smokes?**

If the relationship between BMI and cost is *steeper* for smokers than non-smokers, that would
suggest that smoking and obesity compound each other — meaning a policy targeting both simultaneously
would be more cost-effective than targeting either alone.

**RQ3 — Which modelling approach best predicts individual insurance charges?**

We compare four methods: linear regression, regression trees, random forests, and XGBoost. A more
accurate predictive model helps insurers and health planners identify high-risk individuals earlier,
enabling earlier intervention.

---

## Dataset Description

The dataset contains **1,338 rows** and **7 variables** representing US health insurance
beneficiaries. Each row is one individual. The outcome variable is `charges` — the annual medical
cost billed by the insurer in USD.

| Variable   | Type        | Description                                      |
|------------|-------------|--------------------------------------------------|
| age        | Numeric     | Age of the beneficiary in years                  |
| sex        | Categorical | Male or female                                   |
| bmi        | Numeric     | Body Mass Index (weight/height²)                 |
| children   | Numeric     | Number of dependants covered                     |
| smoker     | Categorical | Whether the person currently smokes              |
| region     | Categorical | US geographic region (NE, NW, SE, SW)            |
| charges    | Numeric     | Annual medical charges billed (USD) — **outcome**|

The dataset contains **no missing values**, so no imputation was required.

---

## Exploratory Analysis

Before building any models, we examine the data to understand its structure and identify any
patterns worth investigating further.

```{r Exploratory Analysis}
str(med_df)
summary(med_df)
colSums(is.na(med_df))
```

### Distribution of charges

```{r charges-distribution}
ggplot(med_df, aes(charges)) +
  geom_histogram(bins = 50, fill = "steelblue", colour = "white") +
  scale_x_continuous(labels = dollar_format()) +
  labs(title = "Distribution of annual insurance charges",
       subtitle = "Strong right skew — a small number of patients incur very high costs",
       x = "Charges (USD)", y = "Count") +
  theme_minimal()
```

The distribution of charges is strongly **right-skewed**: most people incur relatively modest costs,
but a small number of patients generate very high charges. This skew is common in healthcare cost
data and is worth noting because it means the *average* charge is pulled upward by a few extreme
cases — the *median* is a more representative summary of what a typical patient costs.

From a statistical standpoint, this skew also violates the normality assumption that linear
regression relies on for valid p-values and confidence intervals. We test a log-transformation
below to see if it improves the distribution.

```{r log-charges}
ggplot(med_df, aes(log(charges))) +
  geom_histogram(bins = 50, fill = "steelblue", colour = "white") +
  labs(title = "Log-transformed charges",
       subtitle = "Log transformation produces a much more symmetric distribution",
       x = "log(Charges)", y = "Count") +
  theme_minimal()
```

The log-transformed distribution is substantially more symmetric, which is useful to know —
it means if we want reliable inference (trustworthy p-values and confidence intervals) from
linear regression, modelling log(charges) rather than charges directly would be the more
rigorous approach. We retain the original scale here for interpretability, but flag this
as a limitation when reviewing the linear model diagnostics.

---

### Correlations between numeric predictors and charges

```{r corrplot}
med_cor <- med_df |> select(age, bmi, children, charges)
corrplot(cor(med_cor),
         method   = 'number',
         type     = 'lower',
         diag     = FALSE,
         col      = colorRampPalette(c("#D73027", "white", "#1A6099"))(200),
         tl.col   = 'black',
         tl.srt   = 45)
```

Among the numeric variables, **age has the strongest correlation with charges (r = 0.30)**,
followed by BMI (r = 0.20). The number of children is essentially unrelated to cost (r = 0.07).

However, correlation only captures *linear* relationships between two variables at a time —
it does not tell us how variables interact with one another, or how much each predictor
contributes once others are held constant. That requires regression modelling.

---

### Charges by categorical predictors

```{r boxplots}
med_df |>
  pivot_longer(cols = c(smoker, sex, region),
               names_to  = "variable",
               values_to = "category") |>
  ggplot(aes(x = category, y = charges, fill = category)) +
  geom_boxplot(outlier.shape = 21, outlier.size = 1.5, alpha = 0.8) +
  facet_wrap(~ variable, scales = "free_x") +
  scale_y_continuous(labels = dollar_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Annual insurance charges by smoker status, sex, and region",
       subtitle = "Smoking status shows by far the largest difference in median charges",
       x = NULL, y = "Annual charges (USD)") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position    = "none",
    strip.text         = element_text(face = "bold", size = 12),
    strip.background   = element_rect(fill = "grey92", colour = NA),
    axis.text.x        = element_text(angle = 25, hjust = 1),
    panel.grid.major.x = element_blank()
  )
```

Three clear patterns emerge:

- **Smoking** is the most dramatic separator — smokers' median charges are roughly 3× those of
  non-smokers. The spread within the smoker group is also much wider, suggesting other factors
  (likely BMI, as we explore below) create further cost variation *within* smokers.
- **Sex** shows a small difference with males slightly higher, but the distributions overlap
  substantially — sex is unlikely to be a strong independent predictor.
- **Region** shows modest differences, with the southeast slightly higher than other regions,
  possibly reflecting higher obesity rates in that population.

---

### Does the effect of BMI differ by smoking status?

```{r bmi-smoker-scatter}
ggplot(med_df, aes(x = bmi, y = charges, col = smoker)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = 'lm') +
  scale_y_continuous(labels = dollar_format()) +
  scale_colour_manual(values = c("no" = "steelblue", "yes" = "#D73027")) +
  labs(x = "Body Mass Index", y = "Medical charges",
       title = "BMI and charges by smoking status",
       subtitle = "The slope for smokers is much steeper — BMI matters far more if you smoke",
       colour = "Smoker") +
  theme_minimal()
```

This plot directly addresses RQ2. The two trend lines have **very different slopes**:

- For **non-smokers** (blue), the BMI-charges relationship is almost flat — BMI has little
  independent effect on cost for people who do not smoke.
- For **smokers** (red), the slope is steep — each additional BMI unit is associated with
  substantially higher charges.

This divergence is the statistical definition of an **interaction effect**: the impact of one
variable (BMI) depends on the value of another (smoking status). This means any model that treats
BMI and smoking as independent predictors will misrepresent how costs actually accumulate.
We formally test and include this interaction term in the linear regression.

```{r age-smoker-scatter}
ggplot(med_df, aes(x = age, y = charges, col = smoker)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = 'lm') +
  scale_y_continuous(labels = dollar_format()) +
  scale_colour_manual(values = c("no" = "steelblue", "yes" = "#D73027")) +
  labs(title = "Age and charges by smoking status",
       subtitle = "Both groups show increasing costs with age, but slopes are broadly parallel",
       x = "Age (years)", y = "Charges (USD)", colour = "Smoker") +
  theme_minimal()
```

Unlike the BMI plot, the age-charges slopes for smokers and non-smokers are roughly **parallel**
— age increases costs at a similar rate regardless of smoking status. This suggests an age ×
smoking interaction is less warranted. Both groups do show a steady increase with age, consistent
with older patients having more cumulative health needs.

---

## Risk Modelling

We split the data into a **training set (80%, n ≈ 1,070)** used to build models, and a
**test set (20%, n ≈ 268)** held back to evaluate how well each model generalises to new data.
This separation is essential — evaluating a model on the same data it was trained on produces
an overly optimistic picture of real-world performance.

```{r train-test-split}
set.seed(100)
train_index <- createDataPartition(med_df$charges, p = 0.8, list = FALSE)
train <- med_df[train_index, ]
test  <- med_df[-train_index, ]
```

---

### Linear Regression

Linear regression estimates the independent contribution of each predictor to charges, holding
all other variables constant. It produces interpretable coefficients — each coefficient tells us
the expected change in annual charges for a one-unit increase in that variable.

We fit two models: one with main effects only, and one that adds the BMI × smoking interaction
identified in the exploratory analysis.

```{r linear-regression}
lm_model <- lm(charges ~ age + sex + bmi + children + smoker + region,
               data = train)

lm_int_model <- lm(charges ~ age + sex + bmi + children + smoker + region +
                     smoker:bmi,
                   data = train)

tidy(lm_int_model, conf.int = TRUE) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  print()
```

#### Is the interaction term worth including?

```{r anova-test}
anova(lm_model, lm_int_model)
```

The ANOVA F-test compares the two models directly. The result (F = 552, p < 2.2e-16) is
overwhelmingly significant — the interaction term explains a large additional portion of the
variance in charges. **Including it is clearly justified.**

The F-statistic of 552 is very large by any standard; a value above ~4 would typically be
considered significant. A value this large means the improvement from adding the interaction
term is almost certainly not due to chance.

#### Coefficient plot

```{r coef-plot}
tidy(lm_int_model, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  ggplot(aes(x = estimate, y = reorder(term, estimate))) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.3) +
  geom_point(size = 3, colour = "steelblue") +
  scale_x_continuous(labels = dollar_format()) +
  labs(title = "Linear regression coefficients (95% CI)",
       subtitle = "Each bar shows the estimated effect of that variable on annual charges",
       x = "Estimated effect on charges (USD)", y = NULL) +
  theme_minimal()
```

A few things to note from this plot:

- Any predictor whose confidence interval **does not cross zero** (the dashed line) has a
  statistically significant effect at the 5% level.
- **`bmi:smokeryes`** sits far to the right and has a very tight confidence interval — this is
  the interaction term, confirming that each additional BMI unit is worth substantially more
  in charges for smokers than for non-smokers.
- **`smokeryes`** appears negative, which seems counterintuitive. This is a quirk of how
  interaction terms are encoded mathematically: the `smokeryes` coefficient represents the
  effect of smoking when BMI = 0, which is nonsensical. The *true* cost of smoking must be
  calculated at a realistic BMI (see below).
- **Age, BMI, and children** all show positive effects — older patients, higher BMI, and more
  dependants all predict higher costs.
- **Regional effects** are present but modest relative to the clinical predictors.
- **Sex** has a confidence interval close to zero — after controlling for other factors, being
  male adds relatively little to costs independently.

#### What does smoking actually cost at a typical BMI?

```{r smoking-effect}
mean_bmi <- mean(train$bmi)
coefs    <- coef(lm_int_model)

smoking_effect_at_mean_bmi <- coefs["smokeryes"] +
                               coefs["bmi:smokeryes"] * mean_bmi

cat("At the average BMI of", round(mean_bmi, 1),
    "— a smoker is estimated to pay an additional",
    dollar(round(smoking_effect_at_mean_bmi, 0)),
    "per year in insurance charges compared to a non-smoker with the same characteristics.")
```

This is the most policy-relevant number in the linear model. It translates the abstract
coefficient into a concrete dollar figure that a public health planner can act on. An
individual who quits smoking could expect to save roughly this amount annually in medical
costs — and the savings are even larger at higher BMI values.

#### Checking model assumptions

```{r diagnostics, fig.height=8}
autoplot(lm_int_model, which = 1:4, ncol = 2) + theme_minimal()
```

These four diagnostic plots reveal some important limitations of the linear model:

- **Residuals vs Fitted** (top left): The residuals fan outward as fitted values increase — a
  pattern called *heteroscedasticity*. This means the model's errors are larger for high-cost
  patients than low-cost ones. The consequence is that our standard errors, and therefore our
  p-values and confidence intervals, are not fully reliable.
- **Normal Q-Q** (top right): Points in the upper tail deviate from the diagonal line,
  indicating the residuals are more right-skewed than a normal distribution assumes. This
  reinforces the concern above.
- **Scale-Location** (bottom left): The upward trend confirms heteroscedasticity formally.
- **Cook's Distance** (bottom right): A small number of observations (notably #1048 and #1301)
  have high influence on the model's estimates. These are likely patients with unusually high
  charges who are pulling the fitted line toward them.

**The practical takeaway**: the linear model's coefficient estimates are useful for understanding
direction and approximate magnitude of effects, but the exact p-values and confidence intervals
should be interpreted with caution. The tree-based models below do not make these distributional
assumptions and are therefore less sensitive to these issues.

```{r lm-predictions}
lm_pred     <- predict(lm_model, test)
lm_pred_int <- predict(lm_int_model, test)
rmse_lm     <- rmse(test$charges, lm_pred)
rmse_lm_int <- rmse(test$charges, lm_pred_int)
```

---

### Regression Tree

A regression tree makes predictions by splitting the data into groups based on simple yes/no
rules (e.g. "is the person a smoker? is their BMI above 30?"). The result is a flowchart-like
structure that is highly interpretable — you can trace any individual's predicted cost through
the branches. The trade-off is that a single tree tends to be less accurate than more complex
ensemble methods.

The **complexity parameter (cp)** controls how deeply the tree grows. A tree that grows too
deep will memorise the training data but predict poorly on new patients (overfitting). We use
the cp plot to prune to the optimal depth.

```{r regression-tree, fig.height=6}
tree_model <- rpart(charges ~ ., data = train, method = "anova",
                    control = rpart.control(cp = 0.01))

plotcp(tree_model)
```

The cp plot shows prediction error (cross-validated) on the y-axis against tree complexity on
the x-axis. We select the simplest tree whose error is within 1 standard error of the minimum —
the convention is to pick the leftmost point that falls under the horizontal dashed line.

```{r tree-plot, fig.height=7}
rpart.plot(tree_model, type = 4, extra = 101,
           box.palette = "Blues", shadow.col = "grey80")
```

Reading this tree from the top down:

- The **first split is on smoking status** — immediately confirming it as the dominant cost
  driver. This is the single variable that most reduces prediction error when used as the
  first decision rule.
- Among **smokers**, BMI above approximately 30 creates a second tier — obese smokers form
  the highest-cost subgroup in the entire dataset.
- Among **non-smokers**, age becomes the primary differentiator, with older patients predicting
  higher costs — consistent with our earlier correlation and scatter plot findings.

The tree makes the interaction between smoking and BMI tangible: the BMI split only appears in
the smoker branch, not the non-smoker branch, which is a non-parametric confirmation of the
same interaction we detected in the linear model.

```{r tree-rmse}
tree_pred <- predict(tree_model, test)
rmse_tree <- rmse(test$charges, tree_pred)
```

---

### Random Forest

A random forest builds hundreds of individual decision trees, each trained on a random subset
of the data and a random subset of the predictors, then averages their predictions. This
averaging process substantially reduces the variance that makes any single tree unreliable.

We use **10-fold cross-validation** to tune the key hyperparameter `mtry` — the number of
predictors considered at each split. Cross-validation works by repeatedly partitioning the
training data into 10 chunks, training on 9 and validating on the 10th, rotating through
all chunks. This gives a more stable performance estimate than a single split would provide.

```{r random-forest, cache=TRUE}
cv_control <- trainControl(method = "cv", number = 10, savePredictions = "final")

rf_cv <- train(
  charges ~ .,
  data      = train,
  method    = "rf",
  trControl = cv_control,
  tuneGrid  = data.frame(mtry = c(2, 3, 4, 5)),
  ntree     = 500
)

plot(rf_cv)
```

The plot shows cross-validated RMSE (prediction error in USD) for each value of `mtry`. Lower
is better. We select the `mtry` value that minimises CV-RMSE — the model automatically retains
this as the final model.

```{r rf-importance}
rf_model <- rf_cv$finalModel
rf_pred  <- predict(rf_cv, test)
rmse_rf  <- rmse(test$charges, rf_pred)

varImpPlot(rf_model, main = "Random Forest — Variable Importance")
```

Variable importance measures how much prediction accuracy would decrease if each predictor
were removed. **Smoker status and age are the two most important predictors** by a wide margin,
followed by BMI. Sex and region contribute relatively little once the other variables are
accounted for — consistent with our earlier exploratory findings.

---

### XGBoost

XGBoost (Extreme Gradient Boosting) builds trees sequentially, where each new tree focuses on
correcting the errors made by all previous trees. This makes it highly accurate but less
interpretable than a single tree. It is widely used in applied machine learning and often
achieves the best predictive performance on tabular data.

```{r xgboost, cache=TRUE}
x_train <- model.matrix(charges ~ ., data = train)[, -1]
x_test  <- model.matrix(charges ~ ., data = test)[, -1]
y_train <- train$charges

dtrain <- xgb.DMatrix(data = x_train, label = y_train)
dtest  <- xgb.DMatrix(data = x_test,  label = test$charges)

xgb_model <- xgb.train(
  params = list(
    objective        = "reg:squarederror",
    max_depth        = 4,
    eta              = 0.05,
    subsample        = 0.8,
    colsample_bytree = 0.8
  ),
  data                  = dtrain,
  nrounds               = 500,
  watchlist             = list(train = dtrain, eval = dtest),
  early_stopping_rounds = 30,
  print_every_n         = 100,
  verbose               = 1
)

xgb_pred <- predict(xgb_model, dtest)
rmse_xgb <- rmse(test$charges, xgb_pred)
```

```{r xgb-importance}
xgb.importance(model = xgb_model) |>
  xgb.plot.importance(top_n = 10, rel_to_first = TRUE,
                      xlab = "Relative importance")
```

XGBoost's importance rankings largely mirror the random forest's findings: **smoker status,
age, and BMI are the dominant predictors**. The consistency across two fundamentally different
modelling approaches strengthens confidence that these are genuine signals in the data and not
artefacts of one particular method.

---

## Model Comparison

```{r model-comparison}
rsquared <- function(actual, predicted) {
  1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2)
}

results <- data.frame(
  Model = c("Linear Regression", "LM + Interaction",
            "Regression Tree", "Random Forest", "XGBoost"),
  RMSE  = round(c(rmse_lm, rmse_lm_int, rmse_tree, rmse_rf, rmse_xgb), 0),
  R2    = round(c(rsquared(test$charges, lm_pred),
                  rsquared(test$charges, lm_pred_int),
                  rsquared(test$charges, tree_pred),
                  rsquared(test$charges, rf_pred),
                  rsquared(test$charges, xgb_pred)), 3)
) |> arrange(RMSE)

print(results)
```

```{r comparison-plot}
results |>
  ggplot(aes(x = reorder(Model, RMSE), y = RMSE, fill = RMSE)) +
  geom_col() +
  geom_text(aes(label = comma(RMSE)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_gradient(low = "steelblue", high = "#D73027") +
  scale_y_continuous(labels = dollar_format(),
                     expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Model comparison — held-out test set RMSE",
       subtitle = "Lower RMSE means predictions are closer to actual charges on average",
       x = NULL, y = "Root Mean Squared Error (USD)") +
  theme_minimal() +
  theme(legend.position = "none")
```

**RMSE** (Root Mean Squared Error) is the average prediction error in USD — a model with RMSE
of $4,000 is wrong by about $4,000 on average. **R²** is the proportion of variance in charges
explained by the model — an R² of 0.85 means the model accounts for 85% of the variation in
costs across patients.

Key observations:

- Adding the BMI × smoking interaction improves the linear model noticeably, confirming that
  ignoring it was leaving meaningful signal on the table.
- The **random forest and XGBoost substantially outperform** both linear models, indicating
  that the true relationships in this data are not fully captured by straight lines — there
  are non-linearities and interactions that tree-based methods handle automatically.
- The single regression tree sits between the linear models and the ensembles, which is typical:
  it captures non-linearity but its high variance limits accuracy.
- For **prediction**, the ensemble methods are clearly preferred. For **interpretation and policy
  communication**, the linear model with interaction remains valuable because its coefficients
  have direct, communicable meanings.

---

## Partial Dependence Analysis

The model comparison tells us *how accurate* each model is, but not *how* each predictor
influences predictions. Partial dependence plots (PDPs) address this by showing the marginal
effect of one predictor after averaging out all the others — essentially asking "if we held
everything else constant, what would changing this variable do to predicted costs?"

This is more informative than a raw scatter plot, which confounds the effect of each variable
with all others it correlates with.

```{r pdp-bmi, cache=TRUE}
pd_bmi <- partial(rf_cv, pred.var = "bmi", train = train)

pd_bmi |>
  as.data.frame() |>
  ggplot(aes(bmi, yhat)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_vline(xintercept = 25, linetype = "dashed", colour = "#D73027") +
  geom_vline(xintercept = 30, linetype = "dashed", colour = "#D73027") +
  annotate("text", x = 19.5, y = max(pd_bmi$yhat) * 0.97,
           label = "Normal", hjust = 0, colour = "#D73027", size = 3.5) +
  annotate("text", x = 25.5, y = max(pd_bmi$yhat) * 0.97,
           label = "Overweight", hjust = 0, colour = "#D73027", size = 3.5) +
  annotate("text", x = 30.5, y = max(pd_bmi$yhat) * 0.97,
           label = "Obese", hjust = 0, colour = "#D73027", size = 3.5) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title    = "Partial dependence: BMI on predicted charges",
       subtitle = "Costs begin rising noticeably at the overweight threshold (BMI = 25)",
       x = "BMI", y = "Predicted charges (averaged across all other variables)") +
  theme_minimal()
```

The BMI curve is not a straight line — costs are relatively stable in the normal weight range
then begin increasing from the overweight threshold (BMI ≥ 25), accelerating through the obese
range (BMI ≥ 30). This non-linear pattern would not be captured by a linear model without
explicit polynomial terms, but the random forest detects it automatically.

**Policy implication**: weight management interventions that move patients even from the obese
to overweight category — not necessarily to a "healthy" weight — could yield meaningful cost
reductions, suggesting intermediate targets may be achievable and worth pursuing.

```{r pdp-age, cache=TRUE}
pd_age <- partial(rf_cv, pred.var = "age", train = train)

pd_age |>
  as.data.frame() |>
  ggplot(aes(age, yhat)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title    = "Partial dependence: age on predicted charges",
       subtitle = "Costs increase steadily and approximately linearly with age",
       x = "Age (years)", y = "Predicted charges (averaged across all other variables)") +
  theme_minimal()
```

Unlike BMI, the age-charges relationship is approximately linear — costs increase at a steady
rate across all ages. There is no threshold effect or sudden jump. This makes age-based risk
stratification straightforward: older patients are consistently higher cost, and this holds
regardless of their BMI or smoking status (as confirmed by the parallel slopes in the earlier
scatter plot).

Age is not a modifiable risk factor, but it is useful for **risk stratification** — identifying
older patients for preventive screening or early intervention programmes that address the
*modifiable* factors (smoking, BMI) before costs escalate.

```{r pdp-interaction, cache=TRUE}
partial(rf_cv, pred.var = c("bmi", "smoker"), train = train) |>
  as.data.frame() |>
  ggplot(aes(bmi, yhat, colour = smoker)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = dollar_format()) +
  scale_colour_manual(values = c("no" = "steelblue", "yes" = "#D73027")) +
  labs(title    = "Partial dependence: BMI × smoking status",
       subtitle = "The cost penalty for high BMI is far greater in smokers — the two factors compound",
       x = "BMI", y = "Predicted charges", colour = "Smoker") +
  theme_minimal()
```

This is the most important plot in the analysis. It shows directly that:

- For **non-smokers**, BMI has a modest effect on costs across most of the range.
- For **smokers**, higher BMI produces dramatically higher costs, and the gap widens as BMI
  increases.

This interaction means that a smoker with obesity (BMI ≥ 30) does not simply pay the
"smoker penalty" plus the "obesity penalty" as two separate amounts — the combination is
*worse than the sum of the parts*. A public health programme targeting both smoking cessation
and weight management simultaneously in high-BMI smokers would address the highest-cost
subgroup in this dataset.

```{r predicted-vs-actual}
data.frame(
  actual       = test$charges,
  LM           = lm_pred_int,
  RandomForest = rf_pred,
  XGBoost      = xgb_pred
) |>
  pivot_longer(-actual, names_to = "model", values_to = "predicted") |>
  ggplot(aes(x = actual, y = predicted, colour = model)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.4, size = 1.5) +
  facet_wrap(~ model) +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title    = "Predicted vs actual charges — held-out test set",
       subtitle = "Points on the dashed line are perfect predictions. Scatter above/below = error.",
       x = "Actual charges", y = "Predicted charges") +
  theme_minimal() +
  theme(legend.position = "none")
```

These plots reveal *where* each model fails, not just how much:

- The **linear model** shows a clear curved pattern — it systematically underpredicts very high
  costs and overpredicts some low-cost patients. This reflects the non-linearities the model
  cannot capture.
- **Random Forest and XGBoost** track the diagonal more closely across the full range, though
  both still show some scatter at the high end — extreme-cost patients are inherently difficult
  to predict precisely.

---

## Public Health Interpretation

### Summary of findings

Three consistent findings emerge across all modelling approaches, from the simplest correlation
to the most complex ensemble:

**1. Smoking is the single largest modifiable driver of healthcare costs.** Smokers pay
substantially more than non-smokers, and this premium compounds with BMI. At the average BMI
in this dataset, a smoker pays approximately the amount calculated earlier per year in additional
charges. This is not a marginal effect — it is the dominant signal in the data.

**2. The combination of smoking and obesity creates a disproportionately high-cost subgroup.**
This interaction effect appeared in the linear model's formal significance test, in the
regression tree's first two splits, in the random forest's variable importance rankings, and
in the partial dependence plots. Agreement across all four methods means this finding is robust.
Obese smokers are not just the sum of two risk factors — they are a distinct, high-cost group.

**3. Age increases costs steadily but is not interactive with other factors.** Unlike BMI, the
age effect does not depend on smoking status. Age is non-modifiable, but it is a useful tool
for risk stratification — identifying older patients who also smoke or have high BMI, and
prioritising them for preventive intervention, would target resources at the highest-cost
individuals.

### What this means for policy

If the goal is to reduce population-level healthcare costs, this analysis suggests three
prioritised actions:

1. **Smoking cessation programmes** — particularly those targeting people who are also
   overweight or obese. The interaction effect means the cost return on helping a high-BMI
   smoker quit is substantially larger than helping a normal-weight smoker quit.

2. **Weight management support for people who smoke** — even modest reductions in BMI within
   the smoker subgroup are associated with meaningful cost reductions, due to the steep slope
   of the BMI-cost curve for this group.

3. **Age-based risk stratification** — using age as a flag to identify patients for preventive
   screening and behavioural health support before costs escalate. Age alone cannot be addressed,
   but it reliably identifies patients where addressing modifiable factors (smoking, BMI) would
   have the greatest expected impact.

Sex and region had weak independent effects after other factors were controlled for. Policy
resources focused on geography or sex-specific programmes would likely produce smaller returns
than those targeting the behavioural factors above.

### Limitations

- This is a **US insurance dataset** with a specific demographic and coverage structure.
  Direct quantitative translation to Australian public health costs (e.g. NSW Health) would
  require Australian data, though the directional findings are consistent with clinical evidence
  globally.
- The dataset is relatively small (n = 1,338). Confidence intervals around cost estimates are
  correspondingly wide — the exact dollar figures should be treated as indicative rather than
  precise.
- The linear model's assumption violations (heteroscedasticity, non-normality of residuals) mean
  the reported p-values and confidence intervals are approximations. Key conclusions were
  cross-validated against tree-based methods that do not share these assumptions, providing
  additional confidence in the findings.
- This analysis is **observational** — we cannot claim that smoking or obesity *causes* higher
  charges from this data alone, only that they are strongly associated. Causal interpretation
  requires experimental design or longitudinal data with appropriate confounding adjustment.
